---
title: "HDS 5230 High Performance Computing"
subtitle: "Homework Week 4 - PARALLEL CODE, BOOTSTRAPPING, AND PROFILING"
author: "Miao Cai^[Department of Epidemiology and Biostatistics, Saint Louis University. Email address [miao.cai@slu.edu](miao.cai@slu.edu)]"
date: "`r Sys.Date()`"
output: pdf_document
link-citations: yes
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

You will use these datasets to answer some questions listed below. You must be careful to think about what the appropriate denominator is for each question. As you code the answers, be mindful to use the ‘high performance’ coding approaches in data.table.

1) Load up the dataset and convert to a data.table

```{r}
pacman::p_load(gapminder, data.table, parallel, doParallel, foreach, ggplot2)
data(gapminder)
gapminder = data.table(gapminder)
```

2) We will focus on answering the following question: ‘Is the median life expectancy in 2007 for Asia, Americas, Europe, and Africa significantly different from each other? Estimate the median life expectancy in 2007 for Asia, Americas, Europe, and Africa, and bootstrap confidence intervals for each continent.’
  - a. First calculate the median life expectancy for Asia, Americas, Europe, and
Africa for 2007. This is your ‘point estimate’ of the median for each continent.

```{r}
gapminder2007 = gapminder[year == 2007 & continent != 'Oceania']
gapminder2007[, .(median_lifeExp = median(lifeExp,na.rm=T)), keyby = continent]
```

  - b. For the following bootstrapped approach, implement both a parallel version using foreach(), and a non-parallel version using foreach(). Time the difference in the approaches. You can simply use Sys.time() as I did in the lecture. 
  - c. Now bootstrap a distribution of 100,000 possible medians for each continent. You should end up with 4 separate distributions of medians (you must do this separately for each continent – in class, we only did a single bootstrapped distribution).

```{r}
detectCores()
cores = 8
cl <- makeCluster(cores)
registerDoParallel(cl)
n_boot = 1e5
gc()

# Not parallel
starttime1 = Sys.time()
full_result1 <- 
  foreach(i=1:n_boot, .combine=function(x,y)rbindlist(list(x,y))) %do% {
  result1_a = gapminder2007[sample(.N, replace = T),
                                 .(median_lifeExp = median(lifeExp, na.rm=T)), 
                                continent]
  result1_a
}
endtime1 <- Sys.time()
endtime1 - starttime1
#hist(result1)
full_result1[continent == 'Asia', quantile(median_lifeExp, c(0.025, 0.975))]

# Parallel version
rm(result1_a)
gc()

starttime2 <- Sys.time()
full_result2 <- foreach(i=1:cores,
          .combine='c',
          .packages = 'data.table') %dopar% {
    lapply(1:floor(n_boot/cores), function(x){
      gapminder2007[sample(.N, replace = T),
                    .(median_lifeExp = median(lifeExp, na.rm=T)), 
                    keyby = continent]
    })
}
full_result2 = rbindlist(full_result2)
endtime2 <- Sys.time()
endtime2 - starttime2
#hist(result2)
full_result2[continent == 'Asia', quantile(median_lifeExp, c(0.025, 0.975))]

stopCluster(cl)
```

  - d. Summarize the 95% of these 4 distributions by calculating the 2.5% quantile and the 97.5% quantile. This is done separately for each continent (each distribution). You can just directly calculate this using the quantile function in R like so: quantile(sample, c(.975,.025))
  
```{r}
full_result2[continent != 'Oceania', .(CI_left = quantile(median_lifeExp, 0.025),
                 CI_right = quantile(median_lifeExp, 0.975)), 
             by = continent]
```
  
  
  - e. Create a separate histogram for each of the 4 distributions. Stack these distributions on top of each other. This is easily done with ggplot2 and facet_grid if the data is all in one dataframe in a long format with a group indicator. Here is sample code (without the bootstrap part!):

```{r}
ggplot(full_result2, aes(x=median_lifeExp)) + 
  geom_histogram(aes(y=..density..)) + 
  facet_grid(continent~.)
```

