---
title: "Data Table Gapminder Assignment"
author: "Evan Carey"
date: "February 7, 2019"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1) Load data

First we load and examine the dataset, and turn it into a data.table. 

```{r}
## install/load data
if (!require(gapminder)) install.packages('gapminder')
require(gapminder)
data(gapminder)
## turn into data.table
require(data.table)
gapminder_dt <- data.table(gapminder)
## clean up to save space
rm(gapminder)
gc()
```

# I and J subsets

```{r}
## First 30 rows
gapminder_dt[1:30]
## year is 1952
gapminder_dt[year == 1952]
## Africa continent
gapminder_dt[continent == 'Africa']
## 2007 with country and lifeExp column
gapminder_dt[year == 2007,
             list(country, lifeExp)]
```

# Number 3

## A

```{r}
# Which 5 countries have the highest population in 1952? 
gapminder_dt[year == 1952][order(pop)][,tail(.SD)]
# What about 1987? 
gapminder_dt[year == 1987][order(pop)][,tail(.SD)]
# What about 2007?
gapminder_dt[year == 2007][order(pop)][,tail(.SD)]
```

## B

```{r}
# Which 5 countries have the lowest population in 1952? 
gapminder_dt[year == 1952][order(pop)][,head(.SD)]
# What about 1987? 
gapminder_dt[year == 1987][order(pop)][,head(.SD)]
# What about 2007?
gapminder_dt[year == 2007][order(pop)][,head(.SD)]
```

## C

```{r}
# Which 5 countries have the highest life expectancy in 1952? 
gapminder_dt[year == 1952][order(lifeExp)][,tail(.SD)]
# What about 1987? 
gapminder_dt[year == 1987][order(lifeExp)][,tail(.SD)]
# What about 2007?
gapminder_dt[year == 2007][order(lifeExp)][,tail(.SD)]
```

## D

```{r}
# Which 5 countries have the highest life expectancy in 1952? 
gapminder_dt[year == 1952][order(lifeExp)][,head(.SD)]
# What about 1987? 
gapminder_dt[year == 1987][order(lifeExp)][,head(.SD)]
# What about 2007?
gapminder_dt[year == 2007][order(lifeExp)][,head(.SD)]
```

## E

```{r}
# Calculate the average life expectancy by country across all years for only countries in Asia. 
# Which Country in Asia has the highest and lowest average life expectancy?
gapminder_dt[continent == 'Asia',
             list(mean_life_exp=mean(lifeExp)),
             by=country][order(mean_life_exp)][c(1,.N)]
```

## F

```{r}
# Create a new column that is pop*gdpPercap. Which countries have the highest value for this column?
gapminder_dt[,pop_gdp := pop * gdpPercap]
gapminder_dt[order(pop_gdp)][,tail(.SD)]
```

# 4

What is the correlation between gdpPercap and Life expectancy? Using your data.table object, call the correlation function from the j argument (instead of doing it the 'base' R way).

```{r}
## either way works
gapminder_dt[,cor(gdpPercap,lifeExp)]
## include test statistics
gapminder_dt[,cor.test(gdpPercap,lifeExp)]
```

# 5

I want you to create a linear model estimating the slope between year and lifeExp for each country individually, then extract the slopes/country names and save the resulting table. This resulting table should have one row for every country, with two variables (country and slope). It may be helpful to start by just fitting a model for the whole dataset and figuring out how to extract the slope from that model.

```{r}
## linear model for everyone
mod_result <- 
  gapminder_dt[,lm(lifeExp ~ year)]
## extract coefficients
coef(mod_result)
## extract year slope
coef(mod_result)['year']

## visualize this model
library(ggplot2)
ggplot(gapminder_dt,aes(year, lifeExp)) + 
  geom_point() + 
  geom_smooth(method = 'lm')

## Fit linear model for all, extract slope
gapminder_dt[,coef(lm(lifeExp ~ year))['year']]
# do it by country
year_slope_dt <-
  gapminder_dt[,list(year_slope=coef(lm(lifeExp ~ year))['year']),
               by=country]
```


## a

```{r}
# Which 5 countries have the highest average increase in life expectancy over time (biggest slope)
year_slope_dt[order(year_slope)][,tail(.SD)]
```

## b

```{r}
# Which 5 countries have the lowest average increase in life expectancy over time (smallest slope)
year_slope_dt[order(year_slope)][,head(.SD)]
```
